---
title: "Pre-Processing Calls for Service (Police) for Phoenix"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
        theme: simplex 
        self-contained: true
---

```{r setup, include=FALSE}

knitr::opts_chunk$set( echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      fig.width = 12, 
                      fig.height = 10 )

```

<br>

The [Calls for Police Service](https://www.phoenixopendata.com/dataset/calls-for-service) file from the [data portal](https://www.phoenixopendata.com/) contains all citizen-generated dispatched calls for police service. Officer self-initiated calls and non dispatched calls such as calls for general information or calls that are transferred to other departments such as FIRE for response are not included. The Calls for [Service to Fire]() are available at: **INSERT URL FOR FIRE CALLS FOR SERVICE**. The city updates the [file]("https://www.phoenixopendata.com/dataset/64a60154-3b2d-4583-8fb5-6d5e1b469c28/resource/1d536ee6-7ffb-49c3-bffe-5cdd98a3c97e/download/calls-for-service_calls-for-service_callsforservice.csv") at 11am every day and it contains data beginning November 2015 up to 7 days before the posting date. 

As we did with the crime data (**NEED URL TO CRIME PREPROCESSING HERE**), we need to do some pre-processing. Mainly, cleaning up variables, creating basic objects, and incorporating population data to calculate rates.

<br><br>

## Setup and Loading the Data

<br>

### Preparing the environment

First we need to clear the workspace and load the libraries we are going to use.

<br>

```{r}

# clear workspace
rm( list = ls() )


# load libraries
library( dplyr )        # used for wrangling the data
library( tidyr )        # used for wrangling the data
library( openxlsx )     # for opening an excel file
library( here )         # for referencing the local directory
library( tidycensus )   # getting data from the census API
library( zoo )          # to help with interpolation


```

<br>

### Loading the data

Next we want to load the data and do some cleaning. The data can be called directly from the website link, [here]("https://www.phoenixopendata.com/dataset/64a60154-3b2d-4583-8fb5-6d5e1b469c28/resource/1d536ee6-7ffb-49c3-bffe-5cdd98a3c97e/download/calls-for-service_calls-for-service_callsforservice.csv"). 

<br>

```{r, results="hide"}

# get the data
url <- "https://www.phoenixopendata.com/dataset/64a60154-3b2d-4583-8fb5-6d5e1b469c28/resource/1d536ee6-7ffb-49c3-bffe-5cdd98a3c97e/download/calls-for-service_calls-for-service_callsforservice.csv"

url <- "https://www.phoenixopendata.com/dataset/64a60154-3b2d-4583-8fb5-6d5e1b469c28/resource/d53e8f3f-2ce5-4e58-b03d-fe3721c9354e/download/calls-for-service2016-calls-for-servicecallsforsrvc2016.csv"

# assign the data to an object
cfps_data <- read.csv( url, as.is = TRUE, header = TRUE )

# remove cases that are missing data
cfps_data <- na.omit( cfps_data )

# remove duplicate ids
duplicate_ids <- cfps_data$INCIDENT_NUM[ duplicated( cfps_data$INCIDENT_NUM )]
cfps_data <- cfps_data[ !cfps_data$INCIDENT_NUM %in% duplicate_ids, ]
rm( duplicate_ids )

# need to pull off the date off the call received variable


date_vec <- sub( "\\s+.*$", "", cfps_data$CALL_RECEIVED )
date_vec <- strptime( date_vec, format="%m/%d/%Y %H:%M" )

# clean up the dates
date_vec <- strptime( cfps_data$CALL_RECEIVED, format="%m/%d/%Y %H:%M" )

cfps_data$year   <- format( date_vec, format="%Y" )
cfps_data$month  <- format( date_vec, format="%B" )
cfps_data$day365 <- format( date_vec, format="%j" )
cfps_data$week   <- format( date_vec, format="%V" )

# now we want to clean up the times
cfps_data$hour   <- format( date_vec, format = "%H" ) %>% as.numeric()


!!! check the disposition codes for this

!!!HERE: Need to think about how you want to recode these
! the 

# clean up the variable classifying the cases
cfps_data <- 
  cfps_data %>% 
  mutate( call_type = case_when( 
    UCR.CRIME.CATEGORY == "AGGRAVATED ASSAULT" ~ "Assault",
    UCR.CRIME.CATEGORY == "ARSON" ~ "Arson",
    UCR.CRIME.CATEGORY == "BURGLARY" ~ "Burglary",
    UCR.CRIME.CATEGORY == "DRUG OFFENSE" ~ "Drugs",
    UCR.CRIME.CATEGORY == "LARCENY-THEFT" ~ "Theft",
    UCR.CRIME.CATEGORY == "MURDER AND NON-NEGLIGENT MANSLAUGHTER" ~ "Homicide",
    UCR.CRIME.CATEGORY == "MOTOR VEHICLE THEFT" ~ "MV Theft",
    UCR.CRIME.CATEGORY == "RAPE" ~ "Rape",
    UCR.CRIME.CATEGORY == "ROBBERY" ~ "Robbery" ) )

# drop cases from 2015 (these are dropped because the 2015 cases begin in December)
crimeData <- 
  crimeData %>% 
  filter( year != 2015 )

# drop cases for the most recent month (since the data for the current month are incomplete)
crimeData <- crimeData[ ! ( 
  crimeData$month == format( Sys.Date(), format="%B" ) &
    crimeData$year == format( Sys.Date(), format="%Y" ) 
) , ]

```

<br><br>

## Adjusting for Population

Right now, the `crimeData` object is a list of incidents. When we go to aggregate over months or years, we are going to want to adjust based on population differences. In other words, we will want to calculate the rates (not just examine raw counts).

The rates are computed using population data from the [Census Bureau](https://www.census.gov/topics/population.html). The command below uses the Census Bureau API to pull the data. To see more about how the API is used, see the [phoenix-population.R](/Users/jyoung20/ASU Dropbox/Jacob Young/GitHub_repos/OpenCrimPhx/utils/phoenix-population.R) file. As you will see if you look through the [phoenix-population.R](/Users/jyoung20/ASU Dropbox/Jacob Young/GitHub_repos/OpenCrimPhx/utils/phoenix-population.R) file, population data are available for the years 2016-2022 and interpolated for the years 2023-`r format( Sys.Date(), "%Y" )`.

<br>

```{r}

import::here( "phoenixPopDat",
              .from = here::here( "utils/phoenix-population.R" ),
              .character_only = TRUE )

```

<br>

We now have an object, `phoenixPopDat` that is the yearly population for Phoenix. As noted above, the population for 2023-`r format( Sys.Date(), "%Y" )` are just an estimate, as these are not yet available through the Census Bureau's API. Also, the data for 2020 are interpolated as well due to low response rates. The Census Bureau did released a set of experimental estimates for the 2020 1-year ACS, but for ease of analysis here we will just interpolate the data.

<br><br>

## Appending the Geo Data

We can append the geographic data for the data. This piece calls the [crimeDatGeo2016-2024.rds](https://github.com/jacobtnyoung/OpenCrimPhx/blob/main/data/data-geo/crimeDatGeo2016-2024.rds) file that is created using the [crime-geo.R](/Users/jyoung20/ASU Dropbox/Jacob Young/GitHub_repos/OpenCrimPhx/utils/crime-geo.R) script. It uses the Census API to pull coordinate data.

```{r}

# get the file
crimeDatGeo <- readRDS( here( "data/data-geo/crimeDatGeo2016-2024.rds" ) )

# merge the geographic data with the cases
crimeData <- left_join( crimeData, crimeDatGeo, by = "INC.NUMBER" )

```

<br><br>

## Creating Objects

<br>

### Counts

Now that we have the counts pre-processed and the population data, we can build a few objects that we can use for analysis.

<br>

```{r}

# daily count of crimes
crimesByDay <- 
  crimeData %>% 
  select( year, month, day365 ) %>%   
  filter( !is.na( day365 ) ) %>% 
  group_by( year, month, day365 ) %>% 
  summarize( counts = n() ) %>% 
  ungroup() %>% 
  mutate( day.time = seq( 1, length( counts ) ) ) %>% 
  select( counts, day.time ) %>% 
  mutate( days = 
            seq( 
              as.Date( head( strptime( crimeData$OCCURRED.ON, format="%m/%d/%Y %H:%M" ) ) )[1], 
              as.Date( tail( strptime( crimeData$OCCURRED.ON, format="%m/%d/%Y %H:%M" ) ) )[6], 
              length.out = length( counts ) ) ) %>% 
  arrange( day.time )
crimesByDay <- as.data.frame( crimesByDay )

# monthly count of crimes
crimesByMonth <- 
  crimeData %>% 
  select( year, month ) %>%   
  filter( year != 2015 ) %>%  
  filter( !is.na( year ) ) %>% 
  group_by( year, month ) %>% 
  summarize( counts = n() ) %>% 
  spread( year, counts ) %>% 
  arrange( match( month, month.name ) ) %>% 
  select( !month )

# monthly count of crimes by type
crimesByTypeByMonth <-
  crimeData %>% 
  select( year, month, crime.type ) %>%   
  filter( year != 2015 ) %>%  
  filter( !is.na( year ) ) %>% 
  group_by( year, month, crime.type ) %>% 
  summarize( counts = n() ) %>% 
  arrange( match( month, month.name ) ) %>% 
  select( !month )
crimesByTypeByMonth$month <-factor( crimesByTypeByMonth$month,levels = month.name )
crimesByTypeByMonth$month <-factor( month.abb[crimesByTypeByMonth$month],levels = month.abb )

# crimes by year
crimesByYear <- 
  crimeData %>% 
  select( year ) %>%   
  filter( year != 2015 ) %>%  
  filter( !is.na( year ) ) %>% 
  group_by( year ) %>% 
  summarize( counts = n() ) 

# yearly count of crimes by type
crimesByTypeByYear <-
  crimeData %>% 
  select( year, crime.type ) %>%   
  filter( year != 2015 ) %>%  
  filter( !is.na( year ) ) %>% 
  group_by( year, crime.type ) %>% 
  summarize( counts = n() )


```

<br>

### Rates

Now we want to create objects that capture the rates.

<br>

```{r}

# crime rate is calculated as the count of 
# crimes divided by the population size, then multiplied by 100,000

# calculate the crime rate by month
crimeRatesMonth <- as.data.frame( crimesByMonth )

for ( i in 1: dim( crimeRatesMonth )[2] ){
  crimeRatesMonth[,i] <- ( crimeRatesMonth[,i] / phoenixPopDat[i,2] ) * 100000
}

# calculate the crime rate by year
crimeRatesYear <- as.data.frame( crimesByYear )

for ( i in 1: dim( crimeRatesYear )[1] ){
  crimeRatesYear[i,2] <- ( crimeRatesYear[i,2] / phoenixPopDat[i,2] ) * 100000
}

# rename the counts column to be rates
crimeRatesYear <- crimeRatesYear %>%  rename( rates = counts )


# calculate rates for types by month
crimeRatesMonthType <- as.data.frame( crimesByTypeByMonth )

# assign the population data
crimeRatesMonthType <- crimeRatesMonthType %>%
  left_join( phoenixPopDat, by = "year" )

# create the rates
crimeRatesMonthType <- crimeRatesMonthType %>% 
  mutate( rates = ( ( counts / population ) * 100000 ) )


# calculate rates for types by year
crimeRatesYearType <- as.data.frame( crimesByTypeByYear )

# assign the population data
crimeRatesYearType <- crimeRatesYearType %>%
  left_join( phoenixPopDat, by = "year" )

# create the rates
crimeRatesYearType <- crimeRatesYearType %>% 
  mutate( rates = ( ( counts / population ) * 100000 ) )

```

<br><br>

## Saving the File

We can now save the objects as a file that can be referenced for analysis on separate pages. 

<br>

```{r}

# save the objects as .rds files

saveRDS( crimeData,       file = here( "data/crimeData.rds" ) )
saveRDS( crimesByDay,     file = here( "data/crimesByDay.rds" ) )
saveRDS( crimesByMonth,   file = here( "data/crimesByMonth.rds" ) )
saveRDS( crimesByYear,    file = here( "data/crimesByYear.rds" ) )
saveRDS( crimeRatesMonth, file = here( "data/crimeRatesMonth.rds" ) )
saveRDS( crimeRatesYear,  file = here( "data/crimeRatesYear.rds" ) )

# crimes by type
saveRDS( crimesByTypeByMonth, file = here( "data/crimesByTypeByMonth.rds" ) )
saveRDS( crimesByTypeByYear , file = here( "data/crimesByTypeByYear.rds" ) )
saveRDS( crimeRatesMonthType, file = here( "data/crimeRatesMonthType.rds" ) )
saveRDS( crimeRatesYearType , file = here( "data/crimeRatesMonthYear.rds" ) )


```

If you look in the [data](https://jacobtnyoung.github.io/OpenCrimPhx/data) folder for the repository, you will see that these files have been added. When I run this script, I have to push these files to the repository. But, we can now reference them using the `readRDS()` function.

<br><br>

## Next steps...  

Now that the data are pre-processed, we can reference this object when we conduct analyses. Visit the [Crime in Phoenix](https://jacobtnyoung.github.io/OpenCrimPhx/crime.html) page to see these analyses.  

###

<br>

----

<p align="center">
[Back to Open Criminology Phoenix page](https://jacobtnyoung.github.io/OpenCrimPhx/)
</p>

<br>

***Please*** report any needed corrections to the [Issues](https://github.com/jacobtnyoung/OpenCrimPhx/issues/new) page. Thanks!

<br><br>

###### ***Last updated `r format(Sys.time(), '%d %B, %Y')`***