---
title: "Crime and Unemployment"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
        theme: simplex 
        code_folding: hide
---

```{r setup, include=FALSE}

knitr::opts_chunk$set( echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      eval = TRUE,
                      fig.width = 12, 
                      fig.height = 10 )

```

<br>

?

<br><br>

## Setup

To get going, we will load all the libraries we need. The code to generate everything you see here is hidden (to reduce clutter). But, if you want to see “how we get there”, just click the “Show” button on the right. 

Next, we will load the data. These files are available in the [data](https://jacobtnyoung.github.io/OpenCrimPhx/data/data) folder for the repository.


```{r}

# clear workspace
rm( list = ls() )

# load libraries
library( dplyr )    # used for wrangling the data
library( tidyr )    # used for wrangling the data
library( ggplot2 )  # for plotting
library( cowplot )  # for putting the plots together
library( scales )   # for formatting the text
library( forecast ) # for working with time series data
library( here )     # for referencing the local directory


# define the objects
crimeData       <- readRDS( here( "data/crimeData.rds" ) )
crimesByDay     <- readRDS( here( "data/crimesByDay.rds" ) )
crimesByMonth   <- readRDS( here( "data/crimesByMonth.rds" ) )
crimesByYear    <- readRDS( here( "data/crimesByYear.rds" ) )
crimeRatesMonth <- readRDS( here( "data/crimeRatesMonth.rds" ) )
crimeRatesYear  <- readRDS( here( "data/crimeRatesYear.rds" ) )

crimeRatesMonthType  <- readRDS( here( "data/crimeRatesMonthType.rds" ) )

```

<br>

```{r}

# create a time series object for plotting
monthlyCrimeRateByYear <- ts(
  matrix( as.matrix( crimeRatesMonth ), ncol = 1 ), 
  start=c( 2016, 1 ), 
  end=c( as.numeric( tail( names( crimeRatesMonth ), n=1 ) ), 12 ), frequency=12
)

# render the plot
monthlyCrimeRateByYear %>% 
  ggseasonplot(
    year.labels = TRUE,
    continuous = FALSE,
    main = "Plot of Monthly Crime Rate by Years for Phoenix, AZ",
    col = colorRampPalette(c("#f7968f", "#c41104"))( dim( crimeRatesMonth )[2] ) ) + 
  scale_y_continuous( label = comma ) +
  geom_line( size = 1.2 ) +
  theme_minimal() 

```

<br>

Next we need to load the unemployment data.

```{r}

library( blscrapeR )

series_id <- "LAUMT043806000000003"

unemployment_data <- bls_api( 
  series_id, 
  startyear = head( names( crimeRatesMonth ), n=1 ), 
  endyear = tail( names( crimeRatesMonth ), n=1 )
  )

UnemployByMonth <- 
  unemployment_data %>% 
  select( year, periodName, value) %>%
  mutate( month = factor( unemployment_data$periodName,levels = month.name ) ) %>% 
  select( !periodName ) %>% 
  group_by( year, month ) %>% 
  spread( year, value ) %>% 
  ungroup() %>% 
  select( !month )

# create a time series object for plotting
monthlyUnemploymentRateByYear <- ts(
  matrix( as.matrix( UnemployByMonth ), ncol = 1 ), 
  start=c( 2016, 1 ), 
  end=c( as.numeric( tail( names( UnemployByMonth ), n=1 ) ), 12 ), frequency=12
)

# plot the rates
monthlyUnemploymentRateByYear %>% 
  ggseasonplot( 
    year.labels = TRUE,
    continuous = FALSE,
    main = "Plot of Monthly Unemployment Rate by Years for Phoenix, AZ",
    col = colorRampPalette(c("#f7968f", "#c41104"))( dim( UnemployByMonth )[2] ) ) + 
  scale_y_continuous( label = comma ) +
  geom_line( size = 1.2 ) +
  theme_minimal() 

```

Now, we can plot them both together.

```{r}

# render the plot
crime_seasonplot <- monthlyCrimeRateByYear %>% 
  ggseasonplot( 
    year.labels = TRUE,
    continuous = FALSE,
    main = "Plot of Monthly Crime Rate by Years for Phoenix, AZ",
    col = colorRampPalette(c("#f7968f", "#c41104"))( dim( crimeRatesMonth )[2] ) ) + 
  scale_y_continuous( label = comma ) +
  geom_line( size = 1.2 ) +
  theme_minimal() 

unemployment_seasonplot <- monthlyUnemploymentRateByYear %>% 
  ggseasonplot( 
    year.labels = TRUE,
    continuous = FALSE,
    main = "Plot of Monthly Unemployment Rate by Years for Phoenix, AZ",
    col = colorRampPalette(c("#f7968f", "#c41104"))( dim( UnemployByMonth )[2] ) ) + 
  scale_y_continuous( label = comma ) +
  geom_line( size = 1.2 ) +
  theme_minimal() 

combined_plot <- plot_grid( crime_seasonplot, unemployment_seasonplot, ncol = 1 )
print( combined_plot )


```

Plot both together over time.

```{r}

actual_data <- data.frame(
    dsCrime = time( monthlyCrimeRateByYear ),
    yCrime = as.numeric( monthlyCrimeRateByYear ),
    dsUnemp = time( monthlyUnemploymentRateByYear ),
    yUnemp = as.numeric( monthlyUnemploymentRateByYear )
)

crime_plot <- ggplot() +
  geom_line( data = actual_data, aes(x = dsCrime, y = yCrime), color = "grey40") +
  theme_minimal()

unemp_plot <- ggplot() +
  geom_line( data = actual_data, aes(x = dsUnemp, y = yUnemp), color = "grey40") +
  theme_minimal()


combined_plot <- plot_grid( crime_plot, unemp_plot, ncol = 1 )
print( combined_plot )

```

```{r}

library(MTS)

# Combine the time series into a matrix
combined_data <- cbind( monthlyCrimeRateByYear, monthlyUnemploymentRateByYear )
combined_data <- na.omit( combined_data )

# Fit a VARIMA model
varima_model <- VARMA( combined_data, p = 1, q = 1 )  # p and q are the lag orders

# Print the model summary
summary( varima_model )

# Forecasting using the VARIMA model
forecasts <- VARMApred( varima_model, h = 10 )  # h is the forecast horizon
print(forecasts$pred)

```

<br>

<br>

----

<p align="center">
[Back to Open Criminology Phoenix page](https://jacobtnyoung.github.io/OpenCrimPhx/)
</p>

<br>

***Please*** report any needed corrections to the [Issues](https://github.com/jacobtnyoung/OpenCrimPhx/issues/new) page. Thanks!

<br><br>

###### ***Last updated `r format(Sys.time(), '%d %B, %Y')`***